---
title: "Clustering Evangelicals"
author: "Ryan Burge"
date: "April 27, 2016"
output: html_document
---

```{r message=FALSE, warning=FALSE}
library(foreign)
library(car)
library(dplyr)
library(rpart)
library(fpc)
library(ggplot2)
library(foreign)
library(car)
library(dplyr)
library(rpart)
library(fpc)
library(ggplot2)
```

## Are evangelicals are a monolith? 

A lot of outsiders have begun to wonder if evangelicals are truly a monolithic bloc or are there some cracks beginning to form in what we believe an evangelical to be. I wanted to take a look at that by using a technique that is helps to sort a group into similar cases, also known as a cluster. These clusters are created using an algorithm and will help to divide evangelicals up into camps. 

```{r message=FALSE, warning=FALSE}
gss <- read.dta("D:/reltrad.dta", convert.factors = FALSE)
gss <- subset(gss, year >= 2010)
evan <-subset(gss, evangelical == 1)
```

The first thing I have done is used the RELTRAD coding scheme employed by Steensland et al (2000) to create a unified measure of evangelicalism. Here is a link to my github with that syntax for both [Stata and R:](https://github.com/ryanburge/reltrad). In addition I am just using data from the General Social Survey from 2010 to 2014. That is the three most recent waves of the GSS. The subgroup of evangelicals is sufficiently large (about 1500). I am going to do a bunch of cleaning behind the scenes now to get the data in a usable format. To see the full coding and cleaning syntax I have posted it [here.](https://github.com/ryanburge/clustering/blob/master/datacleaning)

```{r message=FALSE, warning=FALSE, echo=FALSE}
evan$literal <- evan$bible
evan$literal <- recode(evan$literal, "1=1; else=0")
evan$attnd <- evan$attend/8
evan$attnd[is.na(evan$attnd)] <-0
evan$male <- evan$sex
evan$male <- recode(evan$male, "1=1; else=0")

evan$atheist <- evan$spkath
evan$atheist <- recode(evan$atheist, "1=1; 2=0; else=0")
evan$atheist2 <- evan$colath
evan$atheist2 <- recode(evan$atheist2, "4=1; 5=0; else=0")
evan$atheist3 <- evan$libath
evan$atheist3 <- recode(evan$atheist3, "2=1; 1=0; else=0")
evan$racist <- evan$spkrac
evan$racist <- recode(evan$racist, "1=1; 2=0; else=0")
evan$racist2 <- evan$colrac
evan$racist2 <- recode(evan$racist2, "4=1; 5=0; else=0")
evan$racist3 <- evan$librac
evan$racist3 <- recode(evan$racist3, "2=1; 1=0; else=0")
evan$mili <- evan$spkmil
evan$mili <- recode(evan$mili, "1=1; 2=0; else=0")
evan$mili2 <- evan$colmil
evan$mili2 <- recode(evan$mili2, "4=1; 5=0; else=0")
evan$mili3 <- evan$libmil
evan$mili3 <- recode(evan$mili3, "2=1; 1=0; else=0")
evan$comm <- evan$spkcom
evan$comm <- recode(evan$comm, "1=1; 2=0; else=0")
evan$comm2 <- evan$colcom
evan$comm2 <- recode(evan$comm2, "5=1; 4=0; else=0")
evan$comm3 <- evan$libcom
evan$comm3 <- recode(evan$comm3, "2=1; 1=0; else=0")
evan$homo <- evan$spkhomo
evan$homo <- recode(evan$homo, "1=1; 2=0; else=0")
evan$homo2 <- evan$colhomo
evan$homo2 <- recode(evan$homo2, "4=1; 5=0; else=0")
evan$homo3 <- evan$libhomo
evan$homo3 <- recode(evan$homo3, "2=1; 1=0; else=0")
evan$tolerance <- evan$atheist + evan$atheist2 + evan$atheist3 + evan$racist + evan$racist2 + evan$racist3 + evan$comm + evan$comm2 + evan$comm3 + evan$mili + evan$mili2 + evan$mili3 + evan$homo + evan$homo2 + evan$homo3
evan$tolerance <- evan$tolerance/15

gss$atheist <- gss$spkath
gss$atheist <- recode(gss$atheist, "1=1; 2=0; else=0")
gss$atheist2 <- gss$colath
gss$atheist2 <- recode(gss$atheist2, "4=1; 5=0; else=0")
gss$atheist3 <- gss$libath
gss$atheist3 <- recode(gss$atheist3, "2=1; 1=0; else=0")
gss$racist <- gss$spkrac
gss$racist <- recode(gss$racist, "1=1; 2=0; else=0")
gss$racist2 <- gss$colrac
gss$racist2 <- recode(gss$racist2, "4=1; 5=0; else=0")
gss$racist3 <- gss$librac
gss$racist3 <- recode(gss$racist3, "2=1; 1=0; else=0")
gss$mili <- gss$spkmil
gss$mili <- recode(gss$mili, "1=1; 2=0; else=0")
gss$mili2 <- gss$colmil
gss$mili2 <- recode(gss$mili2, "4=1; 5=0; else=0")
gss$mili3 <- gss$libmil
gss$mili3 <- recode(gss$mili3, "2=1; 1=0; else=0")
gss$comm <- gss$spkcom
gss$comm <- recode(gss$comm, "1=1; 2=0; else=0")
gss$comm2 <- gss$colcom
gss$comm2 <- recode(gss$comm2, "5=1; 4=0; else=0")
gss$comm3 <- gss$libcom
gss$comm3 <- recode(gss$comm3, "2=1; 1=0; else=0")
gss$homo <- gss$spkhomo
gss$homo <- recode(gss$homo, "1=1; 2=0; else=0")
gss$homo2 <- gss$colhomo
gss$homo2 <- recode(gss$homo2, "4=1; 5=0; else=0")
gss$homo3 <- gss$libhomo
gss$homo3 <- recode(gss$homo3, "2=1; 1=0; else=0")
gss$tolerance <- gss$atheist + gss$atheist2 + gss$atheist3 + gss$racist + gss$racist2 + gss$racist3 + gss$comm + gss$comm2 + gss$comm3 + gss$mili + gss$mili2 + gss$mili3 + gss$homo + gss$homo2 + gss$homo3
gss$tolerance <- gss$tolerance/15


evan$abortion <- evan$abany
evan$abortion <- recode(evan$abortion, "1=1; else=0")
gss$abortion <- gss$abany
gss$abortion <- recode(gss$abortion, "1=1; else=0")

evan$gaymarriage <- evan$marhomo
evan$gaymarriage <- recode(evan$gaymarriage, "1=1; else=0")
gss$gaymarriage <- gss$marhomo
gss$gaymarriage <- recode(gss$gaymarriage, "1=1; else=0")

evan$gaysex <- evan$homosex
evan$gaysex <- recode(evan$gaysex, "4=1; else=0")

gss$gaysex <- gss$homosex
gss$gaysex <- recode(gss$gaysex, "4=1; else=0")


evan$repubid <- evan$partyid
evan$repubid <- recode(evan$repubid, "0=1; 1=2; 2=3; 3=4; 4=5; 5=6; 6=7; 7=0; 8=0; 9=0")
evan$repubid[is.na(evan$repubid)] <-0

gss$repubid <- gss$partyid
gss$repubid <- recode(gss$repubid, "0=1; 1=2; 2=3; 3=4; 4=5; 5=6; 6=7; 7=0; 8=0; 9=0")
gss$repubid[is.na(gss$repubid)] <-0

evan$welfare <- evan$natfare
evan$welfare <- recode(evan$welfare, "8=0; 9=0")
evan$welfare[is.na(evan$welfare)] <-0
evan$welfare <- evan$welfare/3
evan$military <- evan$natarmsy
evan$military <- recode(evan$military, "8=0; 9=0")
evan$military[is.na(evan$military)] <-0
evan$military <- evan$military/3
evan$environment <- evan$natenviy
evan$environment <- evan$natenviy
evan$environment <- recode(evan$environment, "1=3; 3=1; 2=2; else=0")
evan$environment <- evan$environment/3
evan$drugs <- evan$natdrug
evan$drugs <- recode(evan$drugs, "1=3; 3=1; 2=2; else=0")
evan$drugs <- evan$drugs/3

gss$welfare <- gss$natfare
gss$welfare <- recode(gss$welfare, "8=0; 9=0")
gss$welfare[is.na(gss$welfare)] <-0
gss$welfare <- gss$welfare/3
gss$military <- gss$natarmsy
gss$military <- recode(gss$military, "8=0; 9=0")
gss$military[is.na(gss$military)] <-0
gss$military <- gss$military/3
gss$environment <- gss$natenviy
gss$environment <- gss$natenviy
gss$environment <- recode(gss$environment, "1=3; 3=1; 2=2; else=0")
gss$environment <- gss$environment/3
gss$drugs <- gss$natdrug
gss$drugs <- recode(gss$drugs, "1=3; 3=1; 2=2; else=0")
gss$drugs <- gss$drugs/3
```

I'm going to create a dataframe to just do the clustering, called test. Here's the variables that are included. 

```{r message=FALSE, warning=FALSE}
test <- select(evan, age, literal, attnd, male, tolerance, abortion, gaymarriage, gaysex, repubid, welfare, military, environment, drugs, educ)
```

I can't have any missing data for the clustering to work properly. I have a handful of cases where age is missing, as well as education. I am going to use some prediction modeling to make a guess at both the age and education for the missing values. 

```{r message=FALSE, warning=FALSE}
age_fit <- rpart(age ~ educ + attnd + male + tolerance + abortion + gaymarriage + gaysex + repubid + welfare + military + environment + drugs, data = test[!is.na(test$age), ],
method = 'anova')
test$age[is.na(test$age)] <- predict(age_fit, test[is.na(test$age), ])

educ_fit <- rpart(educ ~ age + attnd + male + tolerance + abortion + gaymarriage + gaysex + repubid + welfare + military + environment + drugs, data = test[!is.na(test$age), ],
method = 'anova')
test$educ[is.na(test$educ)] <- predict(educ_fit, test[is.na(test$educ), ])
```

Now, it's time to run the clustering. But how many clusters should I use? 

```{r message=FALSE, warning=FALSE}
wss <- (nrow(test)-1)*sum(apply(test,2,var))
for (i in 2:15) wss[i] <- sum(kmeans(test,
                                     centers=i,iter.max=1000,algorithm="MacQueen")$withinss)

plot(1:15, wss, type="b", xlab="Number of Clusters",
     ylab="Within groups sum of squares")
```

Using the graph it becomes apparent that the jump from 2 to 3 clusters still gives us a good deal of utility, but that moving to four clusters gives us a lot less. Let's stick to three. 

```{r message=FALSE, warning=FALSE}
k<-kmeans(test, centers=3,iter.max=1000,algorithm="MacQueen")
k$size
k$centers
```

So, we have three decently sized centers. Cluster 1 has 577. Cluster 2 has 637. Cluster 358. That's sufficient. And we can see that the clusters have broken down really nicely into three age brackets with Cluster 1 having the average age of 32, Cluster 2's average is 53, and Cluster 3 is 73. Let's visualize those. 

```{r message=FALSE, warning=FALSE}
dp = discrproj(test, k$cluster)
test$clustname<-k$cluster
test$clustname<-factor(test$clustname)
ggplot(test, aes(dp$proj[,1], dp$proj[,2], color=factor(k$cluster)))+geom_point(pch=19,size=2)+ggtitle("Cluster Visualization")

ggplot(test, aes(dp$proj[,1], dp$proj[,2], color=factor(test$literal)))+geom_point(pch=19,size=2)+facet_grid(clustname~.)+ggtitle("How Is Litearlism Dispersed Through the Clusters")
```
